Dimensionality Reduction with PCA
========================================================

```{r}
library(ggplot2)
library(FactoMineR)
library(car)
library(rpart) # for the car90 dataset
library(cluster)
```

Principal Component Analysis
-----
### What is PCA?
Preparations:
* Data must be all be continuous. (The variance of the vector {A,B,C} can't be calculated.)
** Note that in PCA of GWAS data, the genotypes are coded as number of minor alleles (i.e. 0,1,2)
* Data must be centered (i.e. subtract the mean from each variable so that they have mean 0)
* Data must be scaled so that they have the same variance. Because PCA works by finding directions of maximum variance, allowing variables to have different variances can produce biased results. __Note__: The built-in R functions for PCA, `prcomp` and `princomp` __do not__ scale by default. `prcomp` will scale with the option `scale=TRUE`. You shouldn't be using `princomp` at all. 

### Performing PCA
R has many options for performing PCA, including the builtin functions `prcomp` and `princomp`. However, these are old and crufty, and can require you to know what you're doing to get correct results. Infact, `princomp` calculates PCA by performing eigenvalue decomposition on a correlation matrix and can introduce rounding errors into the results. The R help page for `princomp` suggests not using it at all. (`prcomp` uses the more-modern approach of performing singular-value decomposition on the data matrix itself, which avoids that issue entirely).
I prefer an easy to use package, `FactoMineR` for PCA, and will be using it here. The package also includes factor analysis and other related techniques not discussed here (like correspondence analysis).

```{r loaddata}
data(UScereal, package='MASS')
```

```{r cerealPCA}
# PCA can only be done on numeric variables
numericcols <- sapply(UScereal,is.numeric)
# Perform PCA, skip the plots for now
cereal.pca <- PCA(UScereal[numericcols])
```
### Interpreting PCA
#### Eigenvalues
The __eigenvalue__ of a principal component is the amount of variance explained by that component. These are in order, so the first eigenvalue will be the largest and so on. We can retrieve them from `cereal.pca$eig`.

```{r cerealeig}
cereal.pca$eig
```
As we can see, we've gotten 9 PCs from the data, ordered by the amount of variance explained by the component. The first accounts for `r cereal.pca$eig[1,2]`% of the variance in the data. The first three PCs account for `r cereal.pca$eig[3,3]`% of the data. All of the PCs explain all of the variance in the data.
#### How many components should I use?
The eigenvalues of the PCs tell us how many it is useful to retain or examine. But there's not a really good way to objectively get it. 
No rigorous way to choose:
* __Kaiser rule__: Only keep PCs with eigenvalues over 1.
* __Screeplot__: Plot the eigenvalues of PCs in order and look for an elbow (where the line bends into a flat plane, indicating random noise)
```{r}
# A function to make a prettier screeplot than you'd get with default graphics
fmscree <- function(PCAobj) {
  df <- as.data.frame(PCAobj$eig)
  df$component <- seq(1,nrow(df))
  plot <- ggplot(df,aes(x=component,y=eigenvalue)) + geom_point(size=3) + geom_line()
  plot <- plot + geom_hline(yintercept=1,color='red',lty='dotted') 
  print(plot)
  }
```
```{r screereal}
fmscree(cereal.pca)
```
#### Loadings
The __loadings__ describe the relationships of the _variable_ to each of the components. A positive value for a loading indicates that an increased value for that variable results in an increased value for that PC. Variable level information is available in `cereal.pca$var`. The loadings themselves are in `cereal.pca$var$coord`.
```{r cerealloadings}
cereal.pca$var$coord
```
FactoMineR has a convenient way of plotting loadings, by calling `plot` on PCA objects.
```{r plotvariablecereal}
# choix chooses between individuals and variables.
# Its 'choix' because the FactoMineR people are French.
plot(cereal.pca,choix='var')
```
When you look at the plot you can already see a divide between sugary cereals (higher scores on PC2) vs fibery cereals (lower scores on PC2), but its hard to say anything about PC1, since all of the loadings are positive.
#### Contributions
Related to the loadings, the __contributions__ describe how much of a PC is due to a variable. These are available in `cereal.pca$var$contrib`. 
```{r cerealcontrib}
round(cereal.pca$var$contrib,2)
```
These are percents, and every column sums to 100%. If we look at PC1, we see that it's driven by protein, potassium, and calories mostly. Sugars and fat don't really enter into it. PC2, on the other hand is almost all sugars, calories, and fats. Be sure to check the loadings if you want a good understanding of what's going on with your PCs.
#### PC Scores
The __PC score__ describes where on the PC an _observation_ is. This is typically the most useful part of PCA, because it describes where your sample is in the reduced-dimension principal component space. Plotting them is the usual thing to do. If you have useful labels, use them in the plot.
```{r}
# You can plot, just like before with
# plot(cereal.pca,choix='ind')
# but the labels get a little crowded so I'm going to use ggplot
scoredf <- data.frame(PC1=cereal.pca$ind$coord[,1],
                      PC2=cereal.pca$ind$coord[,2],
                      brand=rownames(UScereal))
ggplot(scoredf,aes(x=PC1,y=PC2,label=brand)) + geom_point() + xlim(-5,10)
ggplot(scoredf,aes(x=PC1,y=PC2,label=brand)) + geom_text(size=3,alpha=.75) + xlim(-5,10)
```

With how long the labels are, the graph of the PCs is hard to read. Maybe we'll get some insight on PC1 by looking at the extremes of the distribution.
```{r pc1summary}
summary(scoredf$PC1)
Q1 <- summary(scoredf$PC1)[['1st Qu.']]
Q3 <- summary(scoredf$PC1)[['3rd Qu.']]
# The low scorers
subset(scoredf,PC1 < Q1, select=c(PC1,PC2))

# The high scorers
subset(scoredf,PC1 > Q3, select=c(PC1,PC2))
```
Interpreting PCs is always subjective, but looking at this I think it's clear that PC1 is separating kid's cereal (Fruit Loops, Count Chocula) from grown up cereal (All the bran cereals). An equally valid interpretation is that PC1 is separating candy cereals from healthy cereals.

What about PC2?
```{r pc2summary}
summary(scoredf$PC2)
Q1 <- summary(scoredf$PC2)[['1st Qu.']]
Q3 <- summary(scoredf$PC2)[['3rd Qu.']]
# The low scorers
subset(scoredf,PC2 < Q1, select=c(PC1,PC2))
# The high scorers
subset(scoredf,PC2 > Q3, select=c(PC1,PC2))
```
This one is not as clear as the others, but by looking at this in combination with the loadings for PC2 from the previous section, we can say that that PC2 is separating sugary cereals and fibery cereals.

You can keep digging through for more information from additional PCs, but I couldn't pick out a good pattern from the loadings for PC3 (lower for increased calories but higher for increased sugar?), and since it already has a low eigenvalue (`r cereal.pca$eig[3,1]`), I'll call the rest of the variation noise.

More Examples
------
### Comparing cars by their specs.
The `mtcars` dataset has specs and performance for 32 cars from an issue of Motor Trend from 1972. We're going to use 9 of the predictors (skipping two categorical ones). When we look at the correlations, we see there's a good amount of 
```{r loadmtcars}
data(mtcars)
symnum(cor(mtcars[,-c(8,9)]))
```
Let's see what comes up in the PCA
```{r mtcarspca}
mtcars.pca <- PCA(mtcars[,-c(8,9)],graph=F)
summary(mtcars.pca)
# Make a screeplot to visualize the eigenvalues
fmscree(mtcars.pca)
```
Looking at the Scree plot, its clear that there are only two major principal components, accounting for 85% of the total variance in the dataset. What do they represent?
```{r plotmtcarspcaind}
# choix tells whether to plot individual PC scores or PC loadings for variables
plot(mtcars.pca,choix='ind')
```
Just by eyeballing it, you can see that PC1 (accounting for ~63% of the variance in the dataset) measures the 'fanciness' of the car, as well as it can by specs and performance. In the negative direction you have economy cars, like Hondas and Fiats, as well as a few higher end models which have lower specification. On the other end you have higher end luxury cars: Maseratis, Chryslers, Cadillacs, etc. 


Principal component 2 seems to separate cars based on performance. The Ferrari, Maserati, and the Pantera (a sports car) all have high scores on this PC. On the other hand, you have cars like the Toyota Corona, the Hornet, and especially the Valiant are very poorly performing cars. 

```{r plotmtcarspcavar}
plot(mtcars.pca,choix='var')
```
This plot shows the relationship of the *variables* to the two principal components. Cars that get better mileage are usually economy cars, and mpg shows that. A larger value for mpg results in a lower score on PC1 (the fancy car PC). On the other hand, fancier cars usually have bigger engines, and larger numbers of cylinders, increased displacement, and weight all results in a higher score on PC1. 

Cars that perform poorly on the quarter mile get lower scores on PC2. These are also typically the lower spec'ed cars, so increased values of qsec result in lower scores also get you a lower score on PC1. The opposite is true for number of carbs and horsepower. 
### Outlier detection: The Florida 2000 election.
It can be hard to identify outliers in high dimensional spaces, where there are so many things going on. A great place to look for weird things happening is the 2000  Florida election, where weird things did happen. The dataset `Florida` in package `car` has vote counts for each florida county, for 10 presidential candidates. What happens when you look at the vote counts with PCA.  
```{r floridapca}
data(Florida, package='car')
fl.pca <- PCA(Florida[1:10], graph=F)
plot(fl.pca,choix='ind')
```
Lets look at the first two PCs. So theres an expected group around 0,0. And you can see the large population counties (Dade, Hillsborough, etc) off a little farther. They're going to separate out in any analysis not normalized by population size because their populations are so much larger than the other counties. But the real star of this plot is Volusia county, which is far, far away on PC2 from the rest of the counties. 
```{r floridapcavarplot}
plot(fl.pca,choix='var')
```
When we look at the variable plot, we see that high scores on PC2 are driven by voting for some people I've never heard of (Harris, Phillips, Browne). So that's a big hint that somethings up. Lets compare Volusia to two (geographically) nearby counties, Flagler and Seminole. 

```{r}
# We'll look at the percents too.
Florida.percent <- round(100 * Florida[1:10] / Florida[,'Total'],3)

volusia.neighbors <- c('VOLUSIA','SEMINOLE','FLAGLER','LAKE','PUTNAM','MARION','ORANGE')
Florida[volusia.neighbors,c('GORE','BUSH','PHILLIPS','HARRIS','BROWNE')]
Florida.percent[volusia.neighbors,c('GORE','BUSH','PHILLIPS','HARRIS','BROWNE')]
```
So it looks like the variable causing the issue is too many people voting for Harris than you might expect, given the surrounding counties. For comparison, the median number of votes for Harris per county was `r median(Florida$HARRIS)`.

In 2000, Volusia county's computers had an error that (among other things) gave 9888 erroneous votes to James Harris, the Socialist Workers Party candidate for president. [Here's a New York Times article on it](http://www.nytimes.com/2000/11/10/us/2000-campaign-florida-vote-democrats-tell-problems-polls-across-florida.html?pagewanted=all&src=pm)

What about principal component #3? It still explains 10% of the variance, and has an eigenvalue (slightly) greater than 1. Let's see if there's anything interesting going on there. Let's look at components 1 and 3 together (we already pretty much know what's gone on with PC2 already).
```{r}
plot(fl.pca,axes=c(1,3), choix='ind')
plot(fl.pca,axes=c(1,3), choix='var')
```
This time we get a similar scenario to PC2. This time with two datapoints sticking out, Alachua and Palm Beach. Positive scores on PC2 are driven by votes for McReynolds (the Socialist Party candidate) and Buchanan (a far-right candidate from the Reform party). 
```{r}
alachua.neighbors <- c('ALACHUA','BRADFORD','UNION','LEVY','PUTNAM','CLAY','COLUMBIA','GILCHRIST')
Florida[alachua.neighbors,c('GORE','BUSH','MCREYNOLDS','BUCHANAN')]
Florida.percent[alachua.neighbors,c('GORE','BUSH','MCREYNOLDS','BUCHANAN')]
```
Alachua has alot more votes for McReynolds than the surrounding counties (even if it only comes out to `r Florida.percent['ALACHUA','MCREYNOLDS']`%). It's also the home of the University of Florida. Since he's the socialist candidate, I'm pretty comfortable chalking that up to college students' votes. 
```{r}
# I'm going with similar counties instead of neighbors, because I think St. Lucie is more like Palm Beach than, say, Hendry County
palmbeach.similar <- c('PALM.BEACH','BROWARD','ST.LUCIE','MARTIN','DADE')
Florida[palmbeach.similar,c('GORE','BUSH','MCREYNOLDS','BUCHANAN')]
Florida.percent[palmbeach.similar,c('GORE','BUSH','MCREYNOLDS','BUCHANAN')]
```
Palm Beach on the other had got way more votes for Buchanan than similar counties (even Broward and Dade, which have larger populations). Palm Beach used butterfly ballots which were thought to confuse voters, who intending to vote for Gore voted for Pat Buchanan. This was a major point of contention in tallying votes. [Here's a Wikipedia page on it](http://en.wikipedia.org/wiki/United_States_presidential_election,_2000_Florida_results#Palm_Beach_County.27s_butterfly_ballots).
### Crabs: reducing many size measurements into a one dimensional feature space
The `crabs` dataset is a catalog of measurements of 8 variables on 200 crabs. We're concerned right now with continuous variables, frontal lobe size (FL), rear width (RW), carapace length (CL), carapace width (CW), and body depth (BD). But species (O and M) and sex are available.

```{r loadcrabs}
data(crabs,package='MASS')
head(crabs)
# These are our variables of interest
sizevars <- 4:8
```

Just by looking at the correlations between these variables it's pretty clear what's going on: all the measurements are strongly correlated with each other. Looking at the screeplot confirms it: There is only one principal component for all the size data. 
```{r corcrabs}
round(cor(crabs[sizevars]),2)
plot(crabs[sizevars])
```
Lets do the PCA anyway and take a look around:
```{r pcacrabs}
c.pca <- PCA(crabs[sizevars], graph=F)
summary(c.pca)
```

As we expected, one principal component accounts for 95% of the variance in the data. We've effectively reduced the 5 size dimensions down to one, while losing very little of the information in the original dataset. By bringing sex and species back into the picture, we can see that O species crabs are significantly 'bigger': that is, they have a higher score on PC1 than species B.  
```{r}
c <- crabs
c$PC1 <- c.pca$ind$coord[,1]
ggplot(c,aes(x=PC1,fill=sp)) + geom_density() + facet_grid(sp~sex)
```
### Reducing dimensionality before clustering

```{r}
library(cluster)
data(car90, package='rpart')
# We only want the numeric predictors
car90n <- car90[,sapply(car90,is.numeric)]
```
Just _a priori_, I'll say there are three classes of cars in the dataset. Let's run the classifier.
```{r}
# Calculate the distances
car90n.dist <- daisy(car90n)
car90n.pam <- pam(car90n.dist, k=3)
```
The point of classification is that an observation in class A should be more similar to other points in class A than to points in other classes. A good measure of this is the __silhouette__, which looks at those similarities. A silhouette maximizes at 1, and a negative value indicates that a data point is probably misclassified (it's more similar to points in other classes than its own.) You can also look at average silhouettes in a class as a measure of how well fit that class is. We can plot silhouettes with `plot` on the object returned from `pam`. 
```{r}
plot(car90n.pam, main='Clustering of nontransformed car90n data')
```
The bars on the chart represents the silhoutte for each datapoint. Its obvious that alot of points are not classified well. On the right hand side you can see the sizes and average silhoutte for the cluster. Clusters 2 and 3 do terrible jobs of classifying the data and class 1 isnt so hot either.
So that classification didn't do well. But we already know that cars _do_ come in different classes. That's how they're manufactured. Sports cars are different than sedans. SUVs and vans come on truck frames and should class together alot of the time if you're going by specs. So what's going on?
#### The curse of dimensionality
What's happening is the *curse of dimensionality*, which is when non-intuitive things start happening when dealing with large predictor datasets. This can be a problem for kNN classifiers (like `pam`), in two different ways.
* __Distances__: One of the big problems for nearest-neighbor approaches in high dimensions is that the euclidiean distance formula $$D(p,q)=\sqrt[n]{(q_{1} - p_{1})^{2}  + (q_{2} - p_{2})^{2} + \cdots + (q_{n} - p_{n})^{2}}$$ considers _all_ the features. As you add more features, distances become larger, and you have to go farther and farther to find those nearest neighbors. This can often result in picking a neighbor with the 'wrong' class, and soon your classification system is shot, because pretty much every datapoint has a neighbor in the wrong class.
* __Noise__: The other big problem is that kNN approaches consider all of the predictors when doing the classification. I know this sounds like the last bullet point but it's also bad for a different reason. If the predictors are noisy, that noise is reflected in the the classifications. If you have a predictor that should be completely unrelated to the class, that will be reflected in the classification too.

If the main advantages of PCA are _dimensionality reduction_ (by keeping the PCs that explain a large proportion of variance) and _noise reduction_ (by ignoring the extraneous PCs that don't explain much variance), it looks like it might be something that can help remedy these problems. Lets try projecting the `car90` data down into a lower dimensionality space.

```{r car90npca}
car90n.pca <- PCA(car90n, graph=F, ncp=5)
fmscree(car90n.pca)
```
There are 5 major PCs in the data, going by either the screeplot or the Kaiser rule. I've kept data on the 5 PCs in the output of `PCA` with the `ncp` flag. Let's try clustering *that* data and looking at how the clustering goes.
```{r}
car90n.pca.dist <- daisy(car90n.pca$ind$coord)
car90n.pca.pam <- pam(car90n.pca.dist, k=3)
plot(car90n.pca.pam, main='Clustering of PCs of car90n data')
```
This way is a lot better! The average silhouttes are much higher and theres only one value that looks misclassified. 

### You can get prinicpal components out of nothing.
Principal components can always be found. That doesn't necessarily mean they mean something. Here's some randomly generated data: 200 'observations' of 5 variables, all random normal variates. 
```{r}
summary(PCA(matrix(rnorm(1000),ncol=5)))
```

Special cases and related concepts
------
* Sparse PCA
* Principal component regression

Kernel Methods
-----
### What is a kernel?
